---
title: "Effect of bottom coverage to larval presence"
subtitle: "Week4-ex3, solution"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(StanHeaders)
library(rstan)

library(coda)
set.seed(123)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

Let's first explore the data a bit more.
```{r}
# Read the data
data = read.csv("white_fishes_data.csv")
# Form a data table for sites without bottom vegetation
y.noveg = table(data$AREANAME[data$BOTTOMCOV==0], data$WHIBIN[data$BOTTOMCOV==0])
colnames(y.noveg) <- c("y=0","y=1")
N.noveg = rowSums(y.noveg)
# Form a data table for sites with bottom vegetation
y.veg = table(data$AREANAME[data$BOTTOMCOV==1], data$WHIBIN[data$BOTTOMCOV==1])
colnames(y.veg) <- c("y=0","y=1")
N.veg = rowSums(y.veg)

par(mfrow=c(1,2))
plot(N.veg, main="Number of sampling sites", xlab="Area index", ylab="Number of sites")
points(N.noveg, col="red")
legend(1, 39, c("veg.","no veg."),col=c("black","red"), pch=1, cex=1, box.lty=1)
plot(y.veg[,2]/N.veg, main="proportion of sites with whitefish", xlab="Area index", ylab="proportion")
points(y.noveg[,2]/N.noveg, col="red")
print(y.veg)
```

# Part 1

## The model

```{r}
model="
  data{
   int i;
   int c;
   int <lower = 0> N[i,c];
   int <lower = 0> y[i,c];
   
  }
  
  parameters{
    real <lower = 0, upper = 1> mu[c];
    real <lower = 0> s[c];
    real <lower = 0,upper = 1>  theta[i,c];
  }
  
  model{
    
    for (j in 1:c){
      mu[j] ~ uniform(0,1);
    }
    
    for (j in 1:c){
      s[j] ~ lognormal(4,2);
    }
    
    for(j in 1:i){
      for(k in 1:c){
        theta[j,k] ~ beta( mu[k] * s[k] , s[k] - mu[k] * s[k]);
      }
    }
        
    for(j in 1:i){
      for(k in 1:c){
        y[j,k] ~ binomial(N[j,k],theta[j,k]);
      }
    }
    
  }

"
```

```{r}
dataset <- list("c"=2,"i"=19,"y" = cbind(y.noveg[,2],append(y.veg[,2], 0, after = 6)),"N"=cbind(N.noveg,append(N.veg, 0, after = 6)) )
#give

init1 = list(y=0)
init2 = list(y=10)
init3 = list(y=20)

inits <- list(init1, init2, init3)
# stan function does all of the work of fitting a Stan model and
# returning the results as an instance of stanfit = post in our exercises.
post=stan(model_code=model,data=dataset,init=inits,warmup=500,iter=1000,chains=3,thin=1)
```

### Summary posterior mu

```{r}
summary(post,pars="mu", probs = c(0.025, 0.975))
```

### Summary posterior s

```{r}
summary(post,pars="s", probs = c(0.025, 0.975))
```
### Summary posterior theta

```{r}
summary(post,pars="theta", probs = c(0.025, 0.975))
```

### Autocorrelation mu

```{r}
stan_ac(post,"mu",inc_warmup = FALSE, lags = 25, fun.data = "mean_se")
```


### Autocorrelation Theta


```{r}
stan_ac(post,"theta",inc_warmup = FALSE, lags = 25, fun.data = "mean_se")
```

### Autocorrelation s



```{r}
stan_ac(post,"s",inc_warmup = FALSE, lags = 25, fun.data = "mean_se")
```



#### Visualize the posterior


```{r}
#plot histogram of the posterior of lambda (approximation for density function)
plot(post, plotfun = "hist", pars = "mu",bins=50)
```

```{r}
#plot histogram of the posterior of lambda (approximation for density function)
plot(post, plotfun = "hist", pars = "s",bins=50)
```





```{r}
plot(post, plotfun = "hist", pars = "theta",bins=50)
```

## Part 2

### Posterior $\delta\mu$

```{r}
muSamp=as.matrix(post, pars ="mu")
#muSamp
hist(muSamp[,1]-muSamp[,2])
```

### Posterior $\delta\phi$

```{r}
thetaSamp=as.matrix(post, pars ="theta")
#thetaSamp
for (i in 1:19){
  hist(thetaSamp[,i] - thetaSamp[,i+19])
}
```

## Part 3

### y0 predictions nonvege, y1 predictions vege for N_19 = 20 

```{r}
y0 = rep(0 , dim(thetaSamp)[1])
y1 = rep(0 , dim(thetaSamp)[1])

for ( i in 1:dim(thetaSamp)[1] ){
  y0[i] = rbinom(1,10,thetaSamp[i,19]);
  y1[i] = rbinom(1,10,thetaSamp[i,38]);
}
```

```{r}
hist(y0)
hist(y1)
hist(y0 - y1)
```

## Part 4

### y200 predictions nonvege, y201 predictions vege for N_20 = 10 

```{r}
sSamp=as.matrix(post, pars ="s")
newThetaSamp = rep (0,dim(sSamp)[1])
newThetaSamp2 = rep (0,dim(sSamp)[1])

for (k in 1:dim(sSamp)[1]){
 newThetaSamp[k] = rbeta(1,muSamp[k,1] * sSamp[k,1] , sSamp[k,1]- muSamp[k,1] * sSamp[k,1]) 
}

for (k in 1:dim(sSamp)[1]){
 newThetaSamp2[k] = rbeta(1,muSamp[k,2] * sSamp[k,2] , sSamp[k,2]- muSamp[k,2] * sSamp[k,2]) 
}

y200 = rep(0 , dim(sSamp)[1])
y201 = rep(0 , dim(sSamp)[1])

for ( i in 1:dim(sSamp)[1] ){
  y200[i] = rbinom(1,10,newThetaSamp[i]);
  y201[i] = rbinom(1,10,newThetaSamp2[i]);
}

hist(y200)
hist(y201)
hist(y200 - y201)

#theta[j,k] ~ beta( mu[k] * s[k] , s[k] - mu[k] * s[k]);
```
### Task 5

in the last model we are predicting everything from zero. Starting by s to mu to theta and finally y. Therefore it may contain more error than the one before it. Therefore i will take the model of 2.4 instead.

# Grading

**Total 20 points** Each of the steps provides 4 points from correct answer and 2 points from an answer that is towards the right direction but includes minor mistake (e.g. a bug or typo)


# References

Lari Veneranta, Richard Hudd and Jarno Vanhatalo (2013). Reproduction areas of sea-spawning Coregonids reflect the environment in shallow coastal waters. Marine Ecology Progress Series, 477:231-250. <http://www.int-res.com/abstracts/meps/v477/p231-250/>